---
title: "Differential Peak Analysis"
author: "Kennedy Zinn"
date: "2025-10-15"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
---

```{r}
conditions <- c("R179+T", "R179-T", "WT+T", "WT-T")
replicates <- c("R1", "R2")
sample_names <- unlist(lapply(conditions, function(cond){
  paste0(cond, "-EZH2_", replicates)
}))
```

```{r}
# load qc correlation data
corr.df <- read.table("all_target_bams.plotCorrelation.mat.tab", header = TRUE, check.names = FALSE)
head(corr.df)
```
```{r}
library(dplyr)
cor_no_R3 <- corr.df[!grepl("3$", rownames(corr.df)), !grepl("3$", colnames(corr.df))]
head(cor_no_R3)
```
# Quality Checks (revisit)

## Correlation Plots

```{r}
library(corrplot)
corrplot(as.matrix(corr.df), method = c("shade"), col = rainbow(200), col.lim = c(0,1), addgrid.col = "black")
```

```{r}
corrplot(as.matrix(cor_no_R3), method = c("shade"), col = rainbow(200), col.lim = c(0,1), addgrid.col = "black")
```

## PCA
```{r}
# load PCA qc data
pca <- read.table("all_target_bams.plotPCA.tab", header = TRUE, check.names = FALSE)
head(pca)
```

### PC2 v PC3

#### Including third replicate
```{r}
# plot PC 2 v PC3
pc1 <- as.numeric(pca[1, -c(1,26)])
pc2 <- as.numeric(pca[2, -c(1, 26)])
pc3 <- as.numeric(pca[3, -c(1, 26)])
pca_plot <- data.frame(
  Sample = colnames(pca)[-c(1, 26)],
  PC1 = pc1,
  PC2 = pc2,
  PC3 = pc3
)

#plot pc 2 and 3
library(ggplot2)
ggplot(pca_plot, aes(x = PC2, y = PC3, label = Sample)) +
  geom_text(aes(label = Sample), vjust = -0.8, size = 3) +
  theme_minimal() +
  labs(title = "PCA", 
       x = "PC2", 
       y = "PC3") +
  theme(plot.title = element_text(hjust = 0.5))
```

#### Excluding third replicate
```{r}
pca_no_R3 <- pca_plot[!grepl("3$", pca_plot$Sample),]

#plot pc1 and pc2 for first and second replicates only
library(ggplot2)
ggplot(pca_no_R3, aes(x = PC1, y = PC2, label = Sample)) +
  geom_text(aes(label = Sample), vjust = -0.8, size = 3) +
  theme_minimal() +
  labs(title = "PCA", 
       x = "PC2", 
       y = "PC3") +
  theme(plot.title = element_text(hjust = 0.5))
```

### New PCA without third replicate
```{r}
# new pca sans third replicates
new_pca <- prcomp(cor_no_R3)
```

```{r}
variance_explained <- (new_pca$sdev^2 / sum(new_pca$sdev^2)) * 100

# scree plot
plot(1:length(variance_explained), variance_explained,
     type = "b",
     xlab = "Principal Component",
     ylab = "Variance Explained (%)",
     main = "Scree Plot",
     pch = 19,
     col = "steelblue",
     lwd = 2)
```

```{r}
pca_scores <- as.data.frame(new_pca$x)
pca_scores$Sample <- rownames(pca_scores)

ggplot(pca_scores, aes(x = PC1, y = PC2, label = Sample)) +
  geom_text(aes(label = Sample), vjust = -0.8, size = 3) +
  theme_minimal() +
  labs(title = "PCA: First and Second Replicates",
       x = paste0("PC1 (", round(variance_explained[1], 1), "%)"),
       y = paste0("PC2 (", round(variance_explained[2], 1), "%)")) +
  theme(plot.title = element_text(hjust = 0.5))
```

# DESeq2

```{r}
# create colData
m <- sub("([A-Za-z0-9]+)[+-]T-EZH2_R[1-3]", "\\1", sample_names)
tx <- sub(".*([+-])T-EZH2_R[1-3]", "\\1", sample_names)
rep <- sub(".*([1-3])$", "\\1", sample_names)

coldata <- data.frame(
  sample = sample_names,
  mutant = factor(m, levels = c("WT", "R179")),
  treatment = factor(ifelse(tx=="-", 0,1)),
  replicate = factor(rep),
  row.names = sample_names
)
```

```{r}
# load count matrix
count_matrix <- read.csv("dense_countMat.csv", header = TRUE, check.names = FALSE)
colnames(count_matrix) <- gsub(".target.markdup.sorted.bam", "", colnames(count_matrix))
head(count_matrix)
```

```{r}
# remove low count genes
count_matrix <- count_matrix[,-1]
keep <- which(rowSums(count_matrix) > 5)
count_matrix <- count_matrix[keep,]
```

```{r}
library(DESeq2)
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = coldata,
  design = ~replicate+mutant*treatment
)

dds <- DESeq(dds)
```

## PCA

```{r}
# PCA
library(pcaExplorer)
vsd <- vst(dds, blind = FALSE)
```

```{r}
pca_vsd <- plotPCA(vsd, intgroup = c("replicate", "mutant", "treatment"), returnData = TRUE)
percentVar <- round(100 * attr(pca_vsd, "percentVar"))
ggplot(pca_vsd, aes(PC1, PC2, color=treatment, shape=mutant, label = replicate)) +
  geom_point(size=3) +
  geom_text(aes(label = replicate), vjust = -0.8, size = 3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
```

## Mutation:treatment interaction model

```{r}
res_int <- results(dds, name = "mutantR179.treatment1")
summary(res_int)
```

### treatment

```{r}
res_tx <- results(dds, name = "treatment_1_vs_0")
summary(res_tx)
```

### mutation

```{r}
res_mut <- results(dds, name = "mutant_R179_vs_WT")
summary(res_mut)
```

### MA plot

#### treatment

```{r}
plotMA(res_tx, ylim = c(-2,2))
```

#### mutant

```{r}
plotMA(res_mut, ylim = c(-2,2))
```

## no interaction model

```{r}
library(DESeq2)
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = coldata,
  design = ~replicate+mutant+treatment
)

dds <- DESeq(dds)
```

### treatment

```{r}
res_tx <- results(dds, name = "treatment_1_vs_0")
summary(res_tx)
```

### mutation

```{r}
res_mut <- results(dds, name = "mutant_R179_vs_WT")
summary(res_mut)
```

### MA plot

#### treatment

```{r}
plotMA(res_tx, ylim = c(-2,2))
```

#### mutant

```{r}
plotMA(res_mut, ylim = c(-2,2))
```

## Model with treatment:mutation and replicate:mutation interactions

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = coldata,
  design = ~replicate + replicate:mutant + mutant:treatment + mutant + treatment
)

dds<- DESeq(dds)
resultsNames(dds)
```

### treatment*variant interaction

```{r}
res_mt_int <- results(dds, name = "mutantR179.treatment1")
summary(res_mt_int)
```

### replicate*mutant interaction

```{r}
res_rm_int <- results(dds, name = "replicate2.mutantR179")
summary(res_rm_int)
```

## mutant:replicate model
```{r}
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = coldata,
  design = ~replicate + mutant:replicate + mutant + treatment
)

dds<- DESeq(dds)
resultsNames(dds)
```

### mutant:replicate

```{r}
res_rm_int <- results(dds, name = "replicate2.mutantR179")
summary(res_rm_int)
```

### treatment

```{r}
res_tx <- results(dds, name = "treatment_1_vs_0")
summary(res_tx)
```

```{r}
# compare library sizes
# Total counts per sample
total_counts <- colSums(counts(dds))

# Boxplot
boxplot(total_counts ~ colData(dds)$treatment, 
        xlab = "Treatment", ylab = "Total Counts",
        main = "Library Size by Treatment")
```

### mutation

```{r}
res_mut <- results(dds, name = "mutant_R179_vs_WT")
summary(res_mut)
```

```{r}
# compare library sizes
# Total counts per sample
total_counts <- colSums(counts(dds))

# Boxplot
boxplot(total_counts ~ colData(dds)$mutant, 
        xlab = "Mutation", ylab = "Total Counts",
        main = "Library Size by Mutation")
```

### replicate

```{r}
res_rep <- results(dds, name = "replicate_2_vs_1")
summary(res_rep)
```

```{r}
# compare library sizes
# Total counts per sample
total_counts <- colSums(counts(dds))

# Boxplot
boxplot(total_counts ~ colData(dds)$replicate, 
        xlab = "Replicate", ylab = "Total Counts",
        main = "Library Size by Replicate")
```

# Visualizations

## MA plot

### treatment

```{r}
plotMA(res_tx, ylim = c(-2,2))
```

### mutant

```{r}
plotMA(res_mut, ylim = c(-2,2))
```

## Volcano plots

```{r}
library(tidyr)
volcano <- function(res, title){
  # convert results to df and clean
  df <- as.data.frame(res) %>% 
    filter(!is.na(padj) & !is.na(log2FoldChange))
  
  # categorize significance
  df <- df %>% 
    mutate(
      sig = case_when(
        padj < 0.1 & log2FoldChange > 1 ~ "Up",
        padj < 0.1 & log2FoldChange < -1 ~ "Down",
        padj <0.1 ~ "Significant",
        TRUE ~ "Not Significant"
      )
    )
  
  # count peaks in each sig. category
  counts <- df %>% dplyr::count(sig)
  
  # create volcano plot
  p <- ggplot(df, aes(x = log2FoldChange, y = -log10(padj), color = sig)) +
    geom_point(alpha = 0.6, size = 1.5) +
    scale_color_manual(
      values = c("Up" = "#d62728", "Down" = "#1f77b4", 
                 "Significant" = "#ff7f0e", "NS" = "grey70"),
      breaks = c("Up", "Down", "Significant", "NS"),
      labels = c(
        paste0("Up (n=", counts$n[counts$sig == "Up"], ")"),
        paste0("Down (n=", counts$n[counts$sig == "Down"], ")"),
        paste0("Sig, |LFC|<", 1, " (n=", 
               counts$n[counts$sig == "Significant"], ")"),
        paste0("NS (", counts$n[counts$sig == "NS"], ")")
      )
    ) +
    geom_vline(xintercept = c(-1, 1), 
               linetype = "dashed", color = "black", alpha = 0.5) +
    geom_hline(yintercept = -log10(0.1), 
               linetype = "dashed", color = "black", alpha = 0.5) +
    labs(
      title = title,
      x = "Log2 Fold Change",
      y = "-Log10 Adjusted P-value",
      color = "Classification"
    ) +
    theme_classic(base_size = 12) +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0.5, face = "bold")
    )
  
  return(p)
}
```

### Treatment

```{r}
volcano_tx <- volcano(res_tx, "Treatment Effect")
print(volcano_tx)
```

### Mutant

```{r}
volcano_mut <- volcano(res_mut, "R179 Variant Effect")
print(volcano_mut)
```

### Treatment by variant

```{r}
# Get mutant levels (adjust column name if different)
mutant_levels <- levels(dds$mutant)

# Create a list to store results
res_by_mutant <- list()
volcano_by_mutant <- list()

for (mut_level in mutant_levels) {
  
  # Subset DESeqDataSet
  dds_subset <- dds[, dds$mutant == mut_level]
  
  # Update design formula - remove mutant since it's constant in subset
  design(dds_subset) <- ~ replicate + treatment
  
  # Re-run DESeq on subset
  dds_subset <- DESeq(dds_subset)
  
  # Get treatment effect in this mutant background
  res_by_mutant[[mut_level]] <- results(dds_subset, 
                                        name = "treatment_1_vs_0")
  
  # Create volcano plot
  volcano_by_mutant[[mut_level]] <- volcano(
    res_by_mutant[[mut_level]],
    title = paste("Treatment Effect in", mut_level)
  )
  
  # Display plot
  print(volcano_by_mutant[[mut_level]])
  
  # save plot
    ggsave(paste0("volcano_treatment_in_", mut_level, ".pdf"), 
         volcano_by_mutant[[mut_level]], 
         width = 8, height = 6)
    }
```

### Variant by replicate

```{r}
# Get mutant levels (adjust column name if different)
rep_levels <- levels(dds$replicate)

# Create a list to store results
res_by_rep <- list()
volcano_by_rep <- list()

for (rep_level in rep_levels) {
  
  # Subset DESeqDataSet
  dds_subset <- dds[, dds$replicate == rep_level]
  
  # Update design formula - remove replicate since it's constant in subset
  design(dds_subset) <- ~ mutant + treatment
  
  # Re-run DESeq on subset
  dds_subset <- DESeq(dds_subset)
  
  # Get mutation effect in this replicate background
  res_by_rep[[rep_level]] <- results(dds_subset, 
                                        name = "mutant_R179_vs_WT")
  
  # Create volcano plot
  volcano_by_rep[[rep_level]] <- volcano(
    res_by_rep[[rep_level]],
    title = paste("Variant Effect in", rep_level)
  )
  
  # Display plot
  print(volcano_by_rep[[rep_level]])
  
  # save plot
    ggsave(paste0("volcano_treatment_in_replicate", rep_level, ".pdf"), 
         volcano_by_rep[[rep_level]], 
         width = 8, height = 6)
    }
```

# NEW ATTEMPT WITH DIFFCHIPL

```{r}
library(DiffChIPL)

cpmD <- cpmNorm(count_matrix)
design <- model.matrix( ~ replicate + mutant + treatment, data = coldata)

res <- DiffChIPL(cpmD, design, group0 = treatment)
```

```{r}
fitRlimm3 = res$fitDiffL
rtRlimm3 = res$resDE

id_Rlimma_CPM = rownames(rtRlimm3[which(rtRlimm3$adj.P.Val < 0.05),])
 
rtRlimm3 = rtRlimm3[rownames(cpmD),]
aveE = rtRlimm3$AveExpr
logFC = rtRlimm3$logFC
padj = rtRlimm3$adj.P.Val
plotMAVoc2(mean=aveE, logfc=logFC, adj.P.Val=padj, FC=1, 
          padj=0.05, MA=TRUE,
          title=paste0("Rlimma-CPM \n", "treatment","(padj<0.05)\n", 
                       length(id_Rlimma_CPM), " of ", nrow(rtRlimm3) ))
```

